\documentclass[11pt]{article}
\usepackage[margin=0.9in]{geometry}
\usepackage{geometry}
\usepackage{afterpage}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{amsbsy}
\usepackage{array}
\usepackage{bm}
\usepackage{booktabs}
\usepackage{dcolumn}
\usepackage[strict]{changepage}
\usepackage{etoolbox}
\usepackage{float}
\usepackage{ifthen}
\usepackage{pdflscape}
%\usepackage{lscape}
\usepackage{longtable}
\usepackage[colorlinks]{hyperref}
\usepackage{multicol}
\usepackage{multirow}
\usepackage[sort]{natbib}
\usepackage{pgffor}
\usepackage{rotating}
\usepackage{setspace}
\usepackage{subfig}
\usepackage{tabularx}
\usepackage{titlesec}
\usepackage{url}
\usepackage{verbatim}
\usepackage{xr}
\usepackage{longtable}
\usepackage{rotating}
\usepackage{color}
\usepackage{tikz}
\usepackage{lscape}

\newcommand{\iid}{\stackrel{\mathrm{i.i.d.}}{\sim}}

\makeatletter
  \newcommand\regsz{\@setfontsize\regsz{12pt}{6}}
\makeatother

\makeatletter
  \newcommand\smlsz{\@setfontsize\smlsz{10pt}{6}}
\makeatother

\renewcommand\scriptsize{@setfontsize\scriptsize{5}{7}} 

\newcommand\blfootnote[1]{%
  \begingroup
  \renewcommand\thefootnote{}\footnote{#1}%
  \addtocounter{footnote}{-1}%
  \endgroup
}

\restylefloat{table}

\newcommand{\tab}{\hspace*{2em}}



\title{\vspace{-20mm}\regsz \textbf{Maryclare Griffin} \\ 
\vspace{0mm} \hline \vspace{-2mm}
\begin{center}
\begin{tabular}{lcr}
Assistant Professor & & \\
University of Massachusetts Amherst & \hspace{27mm} & \url{http://maryclare.github.io} \\
Department of Mathematics and Statistics & \hspace{27mm} &  \url{maryclaregri@umass.edu}  \\
1446 Lederle Graduate Research Tower & \hspace{27mm} & \\
Amherst, MA 01003 & & \\
\end{tabular}
\end{center}}
\author{\vspace{-20mm}}
\date{\vspace{-20mm}}

\begin{document}

\maketitle
\thispagestyle{empty}
\pagestyle{empty}

\section*{\regsz Employment}

\begin{itemize}
	\item[] \textbf{Department of Mathematics and Statistics}  \hfill 2019-Present \\
	\textbf{University of Massachusetts Amherst}  \\
	Assistant Professor \vspace{-2mm}
	\item[] \textbf{Center for Applied Mathematics}  \hfill 2018-2019 \\
	\textbf{Cornell University} \\
	Postdoctoral Associate \vspace{-2mm}
%	\item[] \textbf{Economics Research Center}  \hfill 2012-2013 \\
%	\textbf{University of Chicago} \\
%	Research Assistant with Dr. James Heckman \vspace{-2mm}
\end{itemize}

\section*{\regsz Education}

\begin{itemize} 
	\item[] \textbf{University of Washington,} Seattle, WA \hfill 2013-2018 \\ 
	Ph.D., Statistics \\
	Advisor: Peter Hoff\text{, } Thesis Title: ``Model-Based Penalized Regression'' \vspace{-2mm}
	\item[] \textbf{Duke University,} Durham, NC \hfill 2016-2018 \\
	Visiting Ph.D. Student \vspace{-2mm}
	\item[]  \textbf{University of Chicago,} Chicago, IL \hfill 2008-2012 \\
	B.A., Economics with Honors, Statistics\vspace{-2mm}
\end{itemize}

\section*{\regsz Publications and Manuscripts}

\begin{itemize}
	\item[] \textbf{Submitted/In Revision} \vspace{-2mm}
	\item[] Kim D., \underline{Griffin M.}, Nolan D., Kitts J., Gile K. G. ``Bayesian Resolution of Discrepant Self-Reported Network Ties.'' \\
		\tab \emph{Submitted, available on request.} \vspace{-2mm}
	\item[] \underline{Griffin, M.}, \href{https://arxiv.org/abs/2408.01617}{``Review and Demonstration of a Mixture Representation for Simulation from Densities Involving Sums of Powers.''} \\
	\tab \emph{Submitted.} \vspace{-2mm}
	\item[] \underline{Griffin, M.}, Samorodnitsky, G., Matteson, D. S. \href{https://arxiv.org/abs/2011.04168}{``Likelihood Inference for Possibly Non-Stationary Processes via Adaptive Overdifferencing.''} \\
	\tab \emph{Minor revisions requested.} \vspace{-2mm}
	\item[]\textbf{Published}\vspace{-2mm}
		\item[] Winn-Nu\~{n}ez, E. T., \underline{Griffin, M.}, Crawford, L. (2024) \href{https://www.sciencedirect.com/science/article/pii/S0167947323002256}{``A Simple Approach for Local and Global Variable Importance in Nonlinear Regression Models.''} \\
	\tab \emph{Computational Statistics and Data Analysis. 194:1-18} \vspace{-2mm}
	\item[] \underline{Griffin, M.} (2024) \href{https://www.tandfonline.com/doi/abs/10.1080/10618600.2023.2256807}{``Improved Pathwise Coordinate Descent for Power Penalties.''} \\
	\tab \emph{Journal of Computational and Graphical Statistics. 33(1):310-315.} \vspace{-2mm}
	\item[] \underline{Griffin, M.}, Hoff, P. D. (2024) \href{https://www-tandfonline-com.silk.library.umass.edu/doi/full/10.1080/10618600.2023.2233577}{``Structured Shrinkage Priors.''} \\
	\tab \emph{Journal of Computational and Graphical Statistics. 33(1):1-14.} \\
	\tab Selected for the \href{https://www.tandfonline.com/journals/uasa20/collections/EditorsChoiceCollectionUASA}{ASA Editors' Choice Collection}. \vspace{-2mm}
	\item[] Gelsinger, M., \underline{Griffin, M.}, Matteson, D. S., Guinness, J. (2023) \href{https://projecteuclid.org/journals/annals-of-applied-statistics/volume-17/issue-3/Log-Gaussian-Cox-process-modeling-of-large-spatial-lightning-data/10.1214/22-AOAS1708.short}{``Log-Gaussian Cox Process Modeling of Large Spatial Lightning Data using Spectral and Laplace Approximations.''} \\
	\tab \emph{Annals of Applied Statistics. 17(3): 2078-2094.} \vspace{-2mm}
	\item[] Zhang, W., \underline{Griffin, M.}, Matteson, D. S. (2023) \href{https://projecteuclid.org/journals/annals-of-applied-statistics/volume-17/issue-1/Modeling-a-nonlinear-biophysical-trend-followed-by-long-memory-equilibrium/10.1214/22-AOAS1655.short}{``Modeling a Nonlinear Biophysical Trend Followed by Long-Memory Equilibrium with Unknown Change Point.''} \\
	\tab \emph{Annals of Applied Statistics. 17(1):860-880.} \vspace{-2mm}
	\item[] \underline{Griffin, M.}, Hoff, P. D. (2020) \href{https://doi.org/10.1080/10618600.2019.1637749}{``Testing Sparsity-Inducing Penalties.''} \\ 
	\tab \emph{ Journal of Computational and Graphical Statistics. 29(1):128-139.}  \vspace{-2mm}
	\item[] \underline{Griffin, M.}, Hoff, P. D. (2019) \href{https://doi.org/10.1016/j.csda.2019.02.005}{``Lasso ANOVA Decompositions for Matrix and Tensor Data.''} \\
	\tab \emph{ Computational Statistics and Data Analysis. 137:181-194}  \vspace{-2mm}
	\item[] \underline{Griffin, M.}, Gile K. J., Fredriksen-Goldsen K., Handcock M. S., Erosheva E. A. (2018) \href{https://projecteuclid.org/euclid.aoas/1542078044}{``A simulation-based framework for assessing the feasibility of respondent-driven sampling for estimating characteristics in populations of lesbian, gay and bisexual older adults.''} \\
	\tab \emph{Annals of Applied Statistics. 12(4):2252-2278.}  \vspace{-2mm}
	\item[] Holstein, C., \underline{Griffin, M.}, Hong J., Sampson P. (2015) \href{https://pubs.acs.org/doi/abs/10.1021/acs.analchem.5b02082}{``A Statistical Method for Determining and Comparing Limits of Detection of Bioassays.''} \\
	\tab \emph{Analytical Chemistry.} 87(19):9795-9801.\vspace{-2mm}
	 \item[] \textbf{Software} \vspace{-2mm}
	 \item[] \underline{Griffin, M.} (2018) \href{https://cran.r-project.org/web/packages/gnorm/index.html}{\texttt{gnorm}}: Generalized Normal/Exponential Power Distribution for \texttt{R},  version 1.0.0.}

\end{itemize}

\section*{\regsz Funding}

\begin{itemize}  
	\item[] National Science Foundation Division of Undergraduate Education  \hfill 2022-2028\\
	S-STEM Award (\href{https://www.nsf.gov/awardsearch/showAward?AWD_ID=2130262&HistoricalAwards=false}{\#2130262}). Direct \$1,381,459, Indirect \$118,315.	\vspace{-2mm}
	\item[] National Science Foundation Division of Mathematical Sciences  \hfill 2021-2024\\
	Statistics Research Award (\href{https://www.nsf.gov/awardsearch/showAward?AWD_ID=2113079&HistoricalAwards=false}{\#2113079}). Direct \$94,033, Indirect \$55,949. \vspace{-2mm}
	\item[] Mutual Mentoring Team Grant. Direct \$5,502. \hfill 2020\vspace{-2mm}
	\item[] National Science Foundation Graduate Research Fellowship  \hfill 2013-2018\vspace{-2mm}
	\item[] University of Washington, Blalock Fellowship \hfill 2013\vspace{-2mm}
%	\item[] University of Chicago Foreign Language Acquisition Grant 	\hfill June 2010\vspace{-2mm} 
%	\item[] Dean's List, University of Chicago \hfill 2008-2012			   					      	
\end{itemize}


\section*{\regsz Honors and Awards}

\begin{itemize}  
	\item[] College of Natural Sciences Faculty Peer Mentoring Award \hfill 2024\vspace{-2mm}
	\item[] NSF Advance Fellow \hfill 2023-2024\vspace{-2mm}
	\item[] Institute for Social Science Research (ISSR) Scholar \hfill 2021-2022\vspace{-2mm}
	\item[] National Science Foundation Graduate Research Fellowship  \hfill 2013-2018\vspace{-2mm}
	\item[] Frontiers in Forecasting Best Poster Prize \hfill 2018\vspace{-2mm}
	\item[] Women in Statistics and Data Science Conference Travel Award \hfill 2016\vspace{-2mm}
	\item[] Survey Research Methods, Government Statistics, and  \vspace{-2mm}
	\item[] Social Statistics Sections Student Paper Award  \hfill 2015\vspace{-2mm}
	\item[] University of Chicago, Goldberg Award in Economics  \hfill 2012\vspace{-2mm}	
%	\item[] University of Chicago Foreign Language Acquisition Grant 	\hfill June 2010\vspace{-2mm} 
%	\item[] Dean's List, University of Chicago \hfill 2008-2012			   					      	
\end{itemize}

\clearpage 

\section*{\regsz Teaching}

\begin{itemize}
	\item[] \textbf{University of Massachusetts Amherst}$\dagger$  \vspace{-2mm}\blfootnote{$\dagger$ 500-numbered courses are regular upper-undergraduate level, 600-numbered courses are graduate topics courses.}
	\item[] Instructor for STAT 540/STAT 630: Statistical Learning  \hfill F2024\vspace{-2mm}
	\item[] Instructor for STAT 535: Statistical Computing  \hfill F2023\vspace{-2mm}
	\item[] Instructor for STAT 525: Regression Analysis  \hfill S2020, F2020, S2021, S2023\vspace{-2mm}
	\item[] Instructor for STATISTC 697TS: Time Series Analysis and Applications  \hfill S2020, S2022\vspace{-2mm}
	\item[] \textbf{Cornell University}  \vspace{-2mm}
	\item[] Instructor for STSCI4550/ILRST4550/ORIE5550: Applied Time Series Analysis \hfill S2019 \vspace{-2mm}
	\item[] \textbf{University of Washington} \vspace{-2mm}
	\item[] Teaching Assistant for CSS\&S564: Bayesian Statistics  \hfill S2016 \vspace{-2mm}
\end{itemize}

\section*{\regsz Presentations}

\begin{itemize}
	\item[] \textbf{Invited Talks} \vspace{-2mm}
	\item[] Bayesian Generalized Linear Models for Correlated Data with Fewer Latent Variables. \\
	\tab \emph{Western North American Region of The International Biometric Society Conference.} \\
	\tab \emph{Fort Collins, CO; June 2024.}\vspace{-2mm}
	\item[] A Simple Approach for Local and Global Variable Importance in Nonlinear Regression Models. \\
	\tab \emph{New England Statistics Symposium, Storrs CT; June 2024.}\vspace{-2mm}
	\item[] Testing and estimation for sparsity-inducing power penalties. \\
	\tab \emph{Dartmouth College, Hanover NH; February 2024.}\vspace{-2mm}
	\item[] Testing and estimation for sparsity-inducing power penalties. \\
	\tab \emph{Rutgers University, New Brunswick, NJ; November 2023.}\vspace{-2mm}
	\item[] Log-Gaussian Cox Process Modeling of Large Spatial Lightening Data using Spectral and Laplace Approximations. \\
	\tab \emph{The EnviBayes Workshop on Complex Environmental Data,} \\
	\tab \emph{Colorado State University, Fort Collins, CO; September 2023.}\vspace{-2mm}
	\item[] Log-Gaussian Cox Process Modeling of Large Spatial Lightening Data using Spectral and Laplace Approximations. \\
	\tab \emph{Western North American Region of The International Biometric Society Conference.} \\
	\tab \emph{Anchorage, AK; June 2023.}\vspace{-2mm}
	\item[] Log-Gaussian Cox Process Modeling of Large Spatial Lightening Data using Spectral and Laplace Approximations. \\
	\tab \emph{New England Statistics Symposium, Boston, MA; June 2023.}\vspace{-2mm}
	\item[] Estimation of Possibly Non-Stationary Long Memory Processes via Adaptive Overdifferencing. \\
	\tab \emph{Conference on Advances in Time Series Analysis with a Celebration of the} \\
	\tab \emph{70th Birthday and Retirement of Professor Ruey Tsay, Chicago, IL; May 2023.} \vspace{-2mm}
	\item[] An Introduction to Model-Based Penalized Regression. \\
	\tab \emph{Department of Statistics,} \\
	\tab \emph{Colby College, Waterville, ME; April 2023.} \vspace{-2mm}
	\item[] Structured Shrinkage Priors. \\
	\tab \emph{Department of Statistics,} \\
	\tab \emph{University of Connecticut, Storrs, CT; April 2023.} \vspace{-2mm}
	\item[] Likelihood Inference for Possibly Non-Stationary Processes via Adaptive Overdifferencing. \\
	\tab \emph{Department of Mathematics and Statistics,} \\
	\tab \emph{Washington University, Virtual; March 2022.} \vspace{-2mm}
	\item[] Likelihood Inference for Possibly Non-Stationary Processes via Adaptive Overdifferencing. \\
	\tab \emph{Booth School of Business, Econometrics and Statistics Colloquium,} \\
	\tab \emph{University of Chicago, Chicago, IL; October 2021.} \vspace{-2mm}
	\item[] Likelihood Inference for Possibly Non-Stationary Processes via Adaptive Overdifferencing. \\
	\tab \emph{New England Statistics Symposium, Providence, RI; October 2021. }\vspace{-2mm}
	\item[] Bayesian generalized linear models for correlated data with fewer latent variables. \\
	\tab \emph{CMStatistics, Virtual; December 2020. } \vspace{-2mm}
	\item[] Structured Shrinkage Priors. \\
	\tab \emph{Department of Statistical Science,}\\ 
	\tab \emph{University College London, Virtual; November 2020.} \vspace{-2mm}
	\item[] Estimation of Possibly Non-Stationary Long Memory Processes via Adaptive \\
	Overdifferencing. \\
	\tab \emph{Paul H. Chook Department of Information Systems and Statistics,}\\ 
	\tab \emph{Baruch College, Virtual; October 2020.} \vspace{-2mm}
	\item[] Estimation of Possibly Non-Stationary Long Memory Processes via Adaptive \\
	Overdifferencing. \\
	\tab \emph{Department of Mathematics,}\\ 
	\tab \emph{University of Maryland, Virtual; October 2020.} \vspace{-2mm}
	\item[] A review of the computational aspects of penalized regression from a model-based perspective. \\
	\tab \emph{Department of Mathematics and Statistics, Applied Mathematics and Computation} \\
	\tab \emph{Seminar,}\\ 
	\tab \emph{University of Massachusetts Amherst, Amherst, MA; February 2020.} \vspace{-2mm}
	\item[] Estimation for possibly non-stationary long memory processes. \\
	\tab \emph{Department of Biostatistics and Epidemiology,}\\ 
	\tab \emph{University of Massachusetts Amherst, Amherst, MA; January 2020.} \vspace{-2mm}
	\item[] Estimation for possibly non-stationary long memory processes. \\
	\tab \emph{Computational Social Science Institute,}\\ 
	\tab \emph{University of Massachusetts Amherst, Amherst, MA; November 2019.} \vspace{-2mm}
	\item[] An Introduction to Model-Based Penalized Regression. \\
	\tab \emph{Department of Mathematics and Statistics,}\\ 
	\tab \emph{Amherst College, Amherst, MA; November 2019.} \vspace{-2mm}
	\item[] Too Many Predictors, Too Few Responses. \\
	\tab \emph{Research Bytes @ MassMutual,}\\ 
	\tab \emph{Western Mass Statistics and Data Science, Amherst, MA; November 2019.} \vspace{-2mm}
	\item[] Structured Shrinkage Priors. \\
	\tab \emph{Department of Statistics,}\\ 
	\tab \emph{George Mason University, Fairfax, VA; September 2019.} \vspace{-2mm}
	\item[] Model-Based Penalized Regression. \\
	\tab \emph{Department of Mathematics and Statistics,}\\ 
	\tab \emph{University of Massachusetts Amherst, Amherst, MA; December 2018.} \vspace{-2mm}
	\item[] Testing Sparsity-Inducing Penalties. \\
	\tab \emph{Department of Biostatistics and Computational Biology Departmental Colloquia,}\\ 
	\tab \emph{University of Rochester, Rochester, NY; October 2018.} \vspace{-2mm}
	\item[] Beyond the Bayesian Lasso: A Review of Continuous Shrinkage Priors. \\
	\tab \emph{Joint Statistical Meetings, Vancouver, WA; August 2018.} \vspace{-2mm}
	\item[] \textbf{Invited Panels} \vspace{-2mm}
	\item[] Wrapping Up and Moving On: Advice for the Final Stretch of Graduate School. \\
	\tab \emph{Women in Statistics and Data Science Conference, Virtual; October 2021.} \vspace{-2mm}
	\item[] Navigating the World of Data Science. \\
	\tab \emph{Voices of Data Science Conference, Virtual; February 2021.} \vspace{-2mm}
	\item[] Using Data to Inform the ASA's Policy on Sexual Misconduct. \\
	\tab \emph{International Conference on Health Policy Statistics, San Diego, CA; January 2020.} \vspace{-2mm}
	\item[] \textbf{Contributed/Topic Contributed Talks} \vspace{-2mm}
	\item[] Bayesian Generalized Linear Models for Correlated Data with Fewer Latent Variables. \\
	\tab \emph{Joint Statistical Meetings,} \\
	\tab \emph{Portland, OR; August 2024.} \vspace{-2mm}
	\item[] Estimation of Possibly Non-Stationary Long Memory Processes via Adaptive Overdifferencing. \\
	\tab \emph{Joint Statistical Meetings,} \\
	\tab \emph{Toronto, ON; August 2023.} \vspace{-2mm}
	\item[] Generalized Structured Shrinkage Priors Using Correlated Scales. \\
	\tab \emph{Joint Statistical Meetings, Virtual; August 2021.} \vspace{-2mm}
	\item[] Structured Shrinkage Priors. \\
	\tab \emph{Joint Statistical Meetings, Denver, CO; August 2019.} \vspace{-2mm}
	\item[] Long Memory Time Series Methods for the Analysis of Electronic Impedance Sensing Data. \\
	\tab \emph{Western North American Region of The International Biometric Society Conference.} \\
	\tab \emph{Portland, OR; June 2019.} \vspace{-2mm}
	\item[] Structured Shrinkage Priors. \\
	\tab \emph{New England Statistics Symposium, Hartford, CT; May 2019.} \vspace{-2mm}
	\item[] Model-Based Testing of Sparsity Inducing Penalties. \\
	\tab \emph{Western North American Region of The International Biometric Society Conference.} \\
	\tab \emph{Santa Fe, NM; June 2017.} \vspace{-2mm}
	\item[] Using Hierarchical Models to Understand P300-Wave--Based Brain-Computer Interface Performance Among Disabled Adults. \\
	\tab \emph{Joint Statistical Meetings, Chicago, IL; August 2016.} \vspace{-2mm}
	\item[] Assessing Feasibility of Respondent-Driven Sampling Using Pilot Data with an Application to Older Lesbian, Gay, and Bisexual Adults. \\
	\tab \emph{Joint Statistical Meetings, Seattle, WA; August 2015.}  \vspace{-2mm}
	\item[] \textbf{Invited Lightning Session}\vspace{-2mm}
	\item[] Log-Gaussian Cox Process Modeling of Large Spatial Lightning Data using Spectral and Laplace Approximations. \\
	\tab \emph{American Geophysical Union Fall Meeting, Virtual; December 2021.}\vspace{-2mm}
	\item[] \textbf{Poster Presentations}\vspace{-2mm}
	\item[] Testing Sparsity-Inducing Penalties. \\
	\tab \emph{21st Meeting of New Researchers in 
Statistics and Probability.}\\ 
	\tab \emph{Fort Collins, CO, August 2019.} \vspace{-2mm}
	\item[] Pathwise Coordinate Descent for Power Penalized Regression. \\
	\tab \emph{Non-convex Optimization and Deep Learning Workshop.}\\ 
	\tab \emph{Boston, MA, January 2019.} \vspace{-2mm}
	\item[] Testing Sparsity-Inducing Penalties. \\
	\tab \emph{Bridging the Divide, Machine Learning in Medicine Symposium.}\\ 
	\tab \emph{Ithaca, NY, September 2018.} \vspace{-2mm}
	\item[] Testing Sparsity-Inducing Penalties. \\
	\tab \emph{Cornell Day of Statistics.}\\ 
	\tab \emph{Ithaca, NY, September 2018.} \vspace{-2mm}
	\item[] Testing Sparsity-Inducing Penalties. \\
	\tab \emph{Frontiers in Forecasting, Institute for Mathematics and Its Applications.}\\ 
	\tab \emph{Minneapolis, MN, February 2018.} \vspace{-2mm}
	\item[] Sparse, Structured Matrix Estimation via $\ell_1$ Penalization of ANOVA Decomposition. \\
	\tab \emph{Opening Workshop of the 2016-2017 Program on Optimization, SAMSI.}\\ 
	\tab \emph{Research Triangle Park, NC, September 2016.}
\end{itemize}

\section*{\regsz Postdoctoral and Student Advising}

\begin{itemize} 
	\item[] \textbf{Dissertation Advisor} \\ 
	Yilin Zhu \hfill  2024-Present \\
	Ning Duan \hfill  2024-Present\vspace{-2mm}
	\item[] \textbf{Oral Exam Chair} \\ 
	Thomas Robacker \hfill Expected Fall 2024\\
	Ning Duan \hfill Summer 2024\\
	Yilin Zhu \hfill Fall 2023\vspace{-2mm}
	\item[] \textbf{Honors Thesis Chair} \\ 
	Maya Jacob \hfill Fall 2024-Present \\
	Vutjiya Senabunyarithi \hfill Fall 2024-Present \\
	Gabrielle Walczak \hfill Summer 2023-Spring 2024\vspace{-2mm}
	\item[] \textbf{Oral Exam Committee Member} \\ 
	Yun Jiang, Department of Mathematics and Statistics \hfill Spring 2023 \vspace{-2mm}
	\item[] \textbf{Doctoral Dissertation Committee Member} \\
	Gabriel Lewis, Department of Economics \hfill 2022-2024\\
	Nutcha Wattanachit, Department of Biostatistics and Epidemiology \hfill 2022-2023\\
	Dongah Kim, Department of Mathematics and Statistics \hfill 2021-2022\vspace{-2mm}
	\item[] \textbf{Independent Studies} \\ 
	Gabrielle Walczak, UMass Amherst \hfill Fall 2023, Spring 2024\vspace{-2mm}
	\item[] \textbf{Undergraduate Research Mentoring} \\ 
	Natalie Pacheco (S-STEM REU), UMass Amherst \hfill Summer 2024\\
	Trung Do (Department REU), UMass Amherst \hfill Summer 2023\\
	Anh Pham Tran (Department REU), UMass Amherst \hfill Summer 2023\\
	Vutjiya Senabunyarithi (Lee SIP REU), UMass Amherst \hfill Summer 2023\\
	Gabrielle Walczak (NSF REU), UMass Amherst \hfill Summer 2023\\
	Dayishaa Daga (Department REU), Mount Holyoke College \hfill Summer 2021\\
	Thyra Tuttle, UMass Amherst \hfill Spring 2021\vspace{-2mm}
	\item[] \textbf{Assigned Visiting Assistant Professor Research Mentor} \hfill 2020-2024 
\end{itemize}


\section*{\regsz Service and Affiliations}

\begin{itemize} 
	\item[] \textbf{Department of Mathematics and Statistics, University of Massachusetts Amherst} \\ 
	Undergraduate Affairs Committee \hfill 2023-2025\\
	Statistics \& Probability Seminar Planning Committee \hfill 2019-2020, 2021-2022, 2023-2025 \\
	Qualifying Exam Committee Member \hfill Spring 2020-Spring 2022, Spring 2023-Fall 2024  \\ 
	Major Advisor for $\sim$10 Students/Year \hfill Fall 2020-Spring 2022, Spring 2023-Spring 2025 \\
	Anti-Racism Committee \hfill 2021-2025 \\
	Onboarding Committee \hfill 2023-2024 \\
	Statistics Faculty Search Committee (Elected) \hfill Spring 2023, 2023-2024 \\
	Bylaws Committee \hfill 2022-2023\\
	Statistics Graduate Admissions Committee \hfill 2021-2022 \\
	Visiting Assistant Professor Faculty Search Committee (Elected) \hfill 2020-2021 \\
	Five College Statistics Liaison \hfill 2020-2021 \\
	Climate Committee \hfill 2019-2020\vspace{-2mm}
	\item[] \textbf{University of Massachusetts Amherst} \\ 
	Early Career Faculty Interview Committee for Associate Dean for Development \hfill 2021-2022
	Computational Social Science Institute (CSSI ) Steering Committee \hfill 2021-2022\vspace{-2mm}
	\item[] \textbf{Committee of Presidents of Statistical Societies (COPSS)} \\ 
	COPSS Diversity, Equity, and Inclusion (DEI) Task Force \hfill 2020-2021\vspace{-2mm}
	\item[] \textbf{American Statistical Association} \\ 
	Biometrics Section Continuing Education Chair \hfill 2024-2025 \\
	SLDS Section Student Paper Award Committee Member \hfill 2021-2024 \\
	ASA Task Force on Sexual Harassment and Assault \hfill 2018-2019\vspace{-2mm}
	\item[] \textbf{New England Statistics Society} \\ 
	Student Competition Judge \hfill 2023-2024\vspace{-2mm}
	\item[] \textbf{University of Washington} \\
	StatCom Member \hfill 2016-2018 \\
	Graduate Student Representative \hfill 2014-2015 \\
	Graduate and Professional Student Senate Representative \hfill 2013-2014\vspace{-2mm}
\end{itemize}

\begin{itemize} 
	\item[] \textbf{Workshops and Conferences}  \vspace{-2mm}
	\item[] Topic Contributed Session Organizer, ``Innovative Adaptive Statistical Models for Time Series Data'' and ``Statistical Applications and Methods in Public Health: Highlights from Recent Mental, Reproductive, Maternal, and Adolescent Health Research with Policy Relevance'' \hfill   \\
	\tab \emph{Joint Statistical Meetings, Portland, OR; August 2024.}  \vspace{-2mm}
	\item[] Invited Session Organizer, ``Finding and Plugging Leaks in the Pipeline to (Bio)Statistics Careers.'' \hfill   \\
	\tab \emph{Western North American Region of The International Biometric Society Conference,} \\
	\tab \emph{Fort Collins, CO; June 2024.}  \vspace{-2mm}
	\item[] Invited Session Organizer, ``Beyond Independent and Identically Distributed: Models for Non-Standard Data'' \hfill   \\
	\tab \emph{New England Statistics Symposium, Storrs, CT; June 2024.}  \vspace{-2mm}
	\item[] Invited Panel Organizer, ``Women in the Statistics Workforce: Perspectives from Academia and Industry
'' \hfill   \\
	\tab \emph{Women in Statistics and Data Science Conference, Seattle, WA; October 2023.}  \vspace{-2mm}
	\item[] Topic Contributed Session Organizer, ``Bayesian Computation for Streaming, Shapes, and Selection'' \hfill   \\
	\tab \emph{Joint Statistical Meetings, Toronto, ON; August 2023.}  \vspace{-2mm}
	\item[] Invited Session Organizer, ``Advances in Scalable Regression Models for Complex Data.'' \hfill   \\
	\tab \emph{Western North American Region of The International Biometric Society Conference,} \\
	\tab \emph{Anchorage, AK; June 2023.}  \vspace{-2mm}
	\item[] Invited Session Organizer, ``Spatial Statistics in a Changing World,'' ``Innovating in Causal Inference,'' and ``Cutting Edge Statistical Methods for Genetics and Genomics Data'' \hfill   \\
	\tab \emph{New England Statistics Symposium, Boston, MA; June 2023.}  \vspace{-2mm}
	\item[] Invited Session Organizer and Chair, ``Modern Solutions to Pressing Problems.'' \hfill   \\
	\tab \emph{CMStatistics, London, UK (Hybrid); December 2021.}  \vspace{-2mm}
	\item[] Invited Session Organizer, ``Innovations in exact and approximate time series analysis.'' \hfill   \\
	\tab \emph{New England Statistics Symposium, Providence RI (Hybrid); October 2021.}  \vspace{-2mm}
	\item[] Invited Session Organizer, ``Contemporary Mixed Model Methodology and Applications.'' \hfill   \\
	\tab \emph{Joint Statistical Meetings, Philadelphia, PA; August 2020.}  \vspace{-2mm}
	\item[] Invited Session Organizer, ``Modern and Practical Solutions to Difficult High Dimensional Regression Problems.'' \hfill   \\
	\tab \emph{Joint Statistical Meetings, Denver, CO; August 2019.}  \vspace{-2mm}
	\item[] Invited Session Organizer, ``Cutting Edge Methods for Modern Problems in Statistical Genetics and Genomics'' \hfill   \\
	\tab \emph{Western North American Region of The International Biometric Society Conference,} \\
	\tab \emph{Portland, OR; June 2019.}  \vspace{-2mm}
\end{itemize}

\begin{itemize}
	\item[] \textbf{Reviewer}  \\
	\tab Annals of Applied Statistics \\ 
	\tab Applied Stochastic Models in Business and Industry \\
	\tab Bayesian Analysis \\
	\tab Biometrika \\
	\tab Biostatistics \\
	\tab Journal of the American Statistical Association, Theory and Methods \\
	\tab Journal of Computational and Graphical Statistics \\ 
	\tab Journal of Econometrics \\
	\tab Journal of the Royal Statistical Society, Series A \\
	\tab NeuroImage \\
	\tab Observational Studies \\
	\tab The R Journal \\
	\tab Stat \\
	\tab Statistica Sinica \\
	\tab Statistics Surveys \\ 
	\tab Technometrics \\
	\tab NSF 2020-2021, 2023-2024 \vspace{-2mm}
	\item[] \textbf{Associate Editor}  \\
	\tab Journal of Computational and Graphical Statistics \\
	\tab  Data Science in Science \vspace{-2mm}
	\item[] \textbf{Member} \\
	\tab American Statistical Association \\
	\tab  Institute of Mathematical Statistics \\
	\tab Caucus for Women in Statistics \\
	\tab  International Biometric Society 
\end{itemize}


\end{document}